---
title: "Intro to R - 1. Introduction"
author: "Michael Hahsler"
output:
  beamer_presentation: 
    colortheme: beaver
    slide_level: 2
    theme: Boadilla
    toc: yes
    highlight: zenburn
    includes:
      in_header: header.tex
institute: OIT, SMU
subtitle: OIT/SMU Libraries Data Science Workshop Series
fontsize: 10pt
titlegraphic: WorldChangersRBStackedWebOnlyrgb.png
---


# What is R?

## What is R?

```{r, echo=FALSE}
options(digits=3)
options(width= 50)
# for sampling
set.seed(1234)
```


\centerline{\includegraphics[width=3cm]{figs/Rlogo.png}}


- R is ``GNU S''. S is a language for statisticians developed at Bell Laboratories by John Chambers et al.
- R is designed by John Chambers and developed by the R Foundation.
- R is a language and environment for statistical computing and graphics
- R is the de facto standard to develop statistical software
- R implements variety of statistical and graphical techniques (linear and nonlinear
modeling, statistical tests, time series analysis, classification, clustering,
...)

## What is R? (cont.)
R provides

- data handling and storage
- operators for calculations on arrays (matrices)
- a large, coherent, integrated collection of intermediate tools for data analysis
- graphical facilities for data analysis and display
- simple and effective programming language (conditionals, loops, user defined recursive functions)
- extension mechanism with a large collection of packages

## Why R?

* R is open-source and free to use.
* R has a large and active community.
* R is used widely in industry. Microsoft offers commercial solutions.
* R provides state-of-the-art algorithm in 15,000+ extension packages on CRAN (2019).
* R easily interfaces with other environments (C++, Python, Tensor Flow, ...)
* R creates beautiful interactive visualizations (as seen in the New York Times
and The Economist)
* RStudio makes creating reports and dash boards easy. 

## Issue?

```{r, echo = FALSE, out.height= '60%', fig.dim=c(4,4), fig.align='center'}
x <- seq(1, 10, length.out = 100)
plot(x, 2^x, type = "l", xlab = "Learning Effort", ylab = "What you can do", col = "red", axes = FALSE); box()
```
\bigskip

**Reason:** R is confusing! 

- Functional programming
- Vectorization
- Many competing ways to do things (base R, tidyverse, ggplot, grid, Sweave, markdown, ...) and they all can be mixed.


# RStudio and a First R Session

## A first session

Start RStudio and type the code in the gray box into the Console:

```{r}
x <- 1:10
x
y <- x + 1
y
```

Look at the objects in the environment (Environment tab in RStudio)
```{r}
ls()		
```

Ending an R session (Session menu in RStudio)
```{r, eval=FALSE}
q()		
```

## How to get help

R comes with detailed online help

```{r, eval=FALSE}
? ls           # get help on ls
help("ls")     # same as above
?? solve       # keyword search
```

\bigskip

Important information on http://cran.r-project.org/

* Manuals section (read: ``An Introduction to R'').
* Task Views section to find packages.
* Many packages have a vignette (see package page).

Other ways to find information

* Search https://stackoverflow.com/
* Just google

## The R language

* R is a (mostly) functional programming language. 

* Expressions are evaluated, printed and the result is lost unless
assigned with `<-` (_Note:_ Don't use `=` for assignments in R!)

* R is case sensitive! 

* Commands are separated by a line break and rarely by a semi-colon (`;`) 

* Expressions are grouped by braces (`{` and `}`)

* Comments start with a number sign (`#`)

\vfill

### Hint
R and RStudio provide very convenient auto-completion by hitting Tab. 


## Working Directory and Files

R sessions use a working directory to read and write files.

```
getwd()
setwd()
```


\vfill

### RStudio Recommendation 

1. Start with a new R-script file.
2. Save it in the folder you want to work.
3. Go to `Session` and select `Set Working Directory -> To Source File Location`
4. Write all your code into the file (not the Console!) and execute with `CTRL-Enter`.



## Data permanency (a.k.a. Workspace and Global Environment)
\small
During an R session, objects are created and stored in the workspace by name. List objects with:

```{r}
ls()
```


Objects can be removed from the workspace.

```{r}
rm(x)  
ls()
```

Objects can be kept over several sessions by saving the workspace (to file called `.RData`).

### Recommendation
Avoid saving the "Workspace" at the end of each session.
  
_Reason:_ Your sessions will get messy and starting R may slow down if `.RData` gets very big. 
You can also remove the `.RData` file manually.  

# R Basics: Vectors and Subsetting

## Vectors

Vectors are the basic data structure in R. Scalars do not exist! Almost 
all numbers are seen as ``numeric'' (double). 

```{r}
42                      # this is a vector of length one!
```

```{r}
x <- c(10.4, 5.6, 3.1, 6.4, 21.7)   # c combines values 
x

1/x			                        # element-wise division

y <- c(x, 0, 0, 0, 0, 0, x)		     # more combination
y
```

## Vector arithmetic

```{r}
x
y
x+y  # elements of the shorter array are recycled!

sum(x)
length(x)
```

## Sequences
\small

```{r}
s1 <- 1:5	      # sequence of integers
s1

s2 <- seq(-1, 1, by = .2)   # using seq()
s2

rep(s1, times = 2)
rep(s1, each = 2)
```

###
Try "`? seq`" and  "`? rep`"


## Logical vectors

```{r}
x

l <- x > 13   # compare each value in x
l
mode(l)
as.numeric(l)	# coerce l into a numeric vector
```

###
The usual relational operators are available (e.g., 
`<, <=, >, >=, ==, !=, & |`). 
See `?"<"` and `?"&"` (quotation marks are necessary!) 


## Missing Values/Infinity
```{r}
z <- c(1:3,NA)
z

ind <- is.na(z)	    # find missing values
ind

0/0		    # creates a NaN (not a number)

1 + NA

2^5000		    # (close to) infinity
```

###
See `?NA` and `?Inf`



## Character vectors
```{r}
string <- c("Hello", "Ola")
string
```

Combining string vectors
```{r}
paste(string, "World!")	    

labs <- paste(c("X","Y"), 1:10, sep = "")
labs
```

###
See `?paste` 

## Factors

\footnotesize

Used for categorical data. Strings are encoded as numbers with a look-up table.

```{r}
(sex <- c("male", "female", "female", "male", "male"))

(sex <- factor(sex))
```

Look-up table
```{r}
levels(sex)
```

R stores an index into the look-up table
```{r}
as.integer(sex)
```

### Warning
R sometimes converts strings into factors (e.g., for data tables) which can lead to problems. 
To get the strings back use `as.character()`.

## Selecting and modifying subsets
\footnotesize

```{r}
x
# select the first element (index starts with 1!)
x[1]	  
# remove the first element
x[-1]	
# select elements (integer vector)
x[2:4]
# select elements (logical vector)
x[x > 7]	 
# replace elements
x[x > 7] <- NA
x
```


## Selecting and modifying subsets II

```{r}
# using names
fruit <- c(5, 10, 1, 20)
names(fruit) <- c("orange", "banana", "apple", "peach")
fruit
lunch <- fruit[c("apple","orange")]
lunch
```

###
See `?"["` 

# Exercises

## Exercises

1. Create a vector with 10 numbers `(3, 12, 6, -5, 0, 8, 15, 1, -10, 7)` 
 and assign it to `x`.
2. What is the ``data type'' of `x`? How can you find out?
3. Subtract 5 from the 2nd, 4th, 6th, etc. element in `x`. 
4. Compute the sum and the average for `x` (there are functions for that).
5. Reverse the order of the elements in `x`.
6. Find out which numbers in x are negative.
7. Remove all entries with negative numbers from `x`.
8. How long is `x` now (there is a function).
9. Remove `x` from the environment/workspace (session).
10. Create the a vector of strings containing 
"CSE 8001", "CSE 8002", ..., "CSE 8100" using paste.


