---
title: "Intro to R - 2. Objects and Data"
author: "Michael Hahsler"
institute: OIT, SMU
output:
  beamer_presentation:
    colortheme: beaver
    highlight: zenburn
    includes:
      in_header: header.tex
    keep_tex: yes
    slide_level: 2
    theme: Boadilla
    toc: yes
  slidy_presentation:
    highlight: zenburn
subtitle: OIT/SMU Libraries Data Science Workshop Series
fontsize: 9pt
titlegraphic: WorldChangersRBStackedWebOnlyrgb.png
---

      
# Objects and Attributes

## Intrinsic attributes: mode

```{r, echo = FALSE}
options(width = 75)
options(digits = 2)
# for sampling
set.seed(1234)
```


All entities in R are called _objects_. 
Objects have the **intrinsic  attributes** `mode` and `length`.

\vfill

```{r}
x <- c(1.5, 2.6, 3.7)
x
mode(x)

y <- as.character(x)	# coercion with as.<datatype>
y
mode(y)
```

\vfill

###
Modes are types "numeric", "complex", "logical", "character", and "raw".  

## Intrinsic attributes: length

\footnotesize

```{r}
x
length(x)	# length attribute

e <- numeric()
e
length(e)

e[5] <- 12	# implicitly changes length
e

length(e) <- 7	# changing the length explicitly
e
```


## Regular attributes

_Regular attributes_ can be read and set using `attr()` and `attributes()`.

\small
\vfill
```{r}
z <- 1:4
z

attributes(z)   # z does not have any attributes

length(z)
class(z)  
mode(z)
storage.mode(z)   
```


## Regular attributes II

Setting an attribute can change the object.
For example, the dim attribute allows R to treat z as a matrix.
\footnotesize
\vfill

```{r}
attr(z, "dim") <- c(2,2)
z

attributes(z)   # returns attributes as a list

length(z)
class(z)        # note: the class has changed!
mode(z)
storage.mode(z)   
```

## Regular attributes III

**Notes:** 

* `mode`, `storage.mode`, and `class` are confusing!

* `class()`returns the class of an object (same as `attr(x, 'class')`). 
If the object does not have
a class attribute (S3) or an implicit class (matrix, array, integer)
then it returns the storage type (`storage.mode`).

\bigskip
<!-- 
### Recommendation
--> 

\begin{block}{Recommendation}
Use `str()` to inspect an object's structure and attributes instead.
\end{block}

Example:

```{r}
str(z)
```


<!--

# Arrays

## Arrays
\footnotesize
Arrays are just vectors with a dim attribute.

```{r}
x <- 1:8
dim(x) <- c(2,2,2)  # convenient accessor for dim attribute
x

x <- array(1:8, dim=c(2,2,2))	# creation using array()

x[2,2,2]	    # access a single element	
```
-->

# Matrices

## Matrix: 2-dimensional array with consistent data type
\footnotesize

```{r}
x <- matrix(1:6, nrow = 2, ncol = 3) # create a matrix
x


x[2, 2] # get one element
x[1, ]	# get the first row
x[ , 2]	# get 2nd column

dim(x)	# look at the dimensions
nrow(x)
ncol(x)

length(x) # intrinsic length attribute (matrix has 6 values) 
```


## Matrix: Dimnames

```{r}
colnames(x) <- c("X1", "X2", "X3")
x
rownames(x) <- c("Michael", "peter")
x
dimnames(x) # names for all dimensions as a list
```

## Matrix: Row and Column operations
\footnotesize
```{r}
x

rowSums(x)
colSums(x)

rowMeans(x)
colMeans(x)

apply(x, MARGIN = 1, FUN = mean) # same as rowMeans(x)
```

\vfill
###
`apply` applies any function to rows (`MARGIN = 1`) or columns (`MARGIN = 2`) 
of a matrix.


## Matrix: rbind, cbind

Construct a matrix by adding another matrix as new columns or rows.

```{r}
m1 <- matrix(TRUE, nrow = 2, ncol = 2)
m0 <- matrix(FALSE, nrow = 2, ncol = 2)

x <- cbind(m0, m1) # binding columns
x

x <- rbind(x, cbind(m1, m0)) # binding rows
x
```

## Matrix algebra (advanced knowledge)
\footnotesize

```{r}
a <- 1:3; b <- 3:1
ab <- outer(a, b, "*")    # outer product
ab

t(ab)			# transpose of a

ab * ab			# element by element product

ab %*% ab		# matrix product
```

\vfill
###
Other important functions: `crossprod()`, 
`solve()` (linear equations), `svd()`, `eigen()`


# Lists

## List

Lists are very common in R.
A list is an object consisting of an ordered collection of objects (its components). 

\tiny
\vfill

```{r}
lst <- list(name = "Fred", wife = "Mary", no.children = 3,
                   child.ages = c(4, 7, 9))

lst

lst[[2]]	    # access via index
lst$wife	    # access via name, also lst[["wife"]]

str(lst)
```

\vfill
###
Lists can contain arbitrary R objects and can be combined with `c()`.
Names can be retrieved and changed with `names()`. 

# Data Frames

## Data Frame: The _spread sheet_ of R

A data frame looks like a spread sheet. It is a list 
of column vectors with class `data.frame`.

\tiny
\vfill

```{r}
df <- data.frame(name = c("Michael", "Mark", "Maggie"), children = c(2, 0, 2))
df

# looks like a list of columns
df$name	    

# also looks like a matrix
df[1, ]	    
df[ , "children"]

str(df)
```

\small
\vfill
### Hints
1. Data structures can be inspected using the Environment tab in RStudio.
2. A data frame is a list of columns and can be accessed like a list. 
3. Character strings are often automatically converted to `factor`.

# S3 Objects

## S3 Objects

S3 objects are just regular R objects with a `class` attribute. Often it is a list.

\vfill
\footnotesize

```{r}
# roll a die 100 times and tabulate the results
dice_rolls <- sample(1:6, size = 100, replace = TRUE)
tbl <- table(dice_rolls)
tbl

attributes(tbl)
```

\vfill

`str()` is very helpful and shows the class.

```{r}
str(tbl)
```


# Importing Data in R

## Accessing R data sets

R (or packages) come with data sets. These sets can be loaded using 
`data()`. Without arguments `data()` shows all available data sets.

\footnotesize
\vfill

```{r}
data(iris)	# load the iris data set. try: ?iris
head(iris)	# head shows the first few elements

str(iris)	# displays the structure of an object
summary(iris)
```


## Reading and writing CSV and text files

\small
`read.table()` and `write.table()` can be used to read/write complete file  to/from data.frames.
The file format can be space or tab-separated, CSV, with our without 
column/row labels, etc. For CSV we have a convenience functions `read.csv()` and `write.csv()`
\vfill


```{r}
df

write.csv(df, file = "df.csv")
df2 <- read.csv("df.csv")
df2

unlink("df.csv")	# remove the file
```

### Notes
1. Don't forget to set the working directory (RStudio: `Session -> Set Working Directory`)
2. See `? read.table` and `? write.table` for all available options (column headers, etc.).

## Importing Excel Files
\footnotesize


__Note:__ You need to 

* Install package `xlsx` using `Tools -> Install Packages...`.
* download the file [MLB_cleaned.xlsx](https://michael.hahsler.net/SMU/DS_Workshop_Intro_R/slides/MLB_cleaned.xlsx)

\vfill

```{r}
library(xlsx)
mlb <- xlsx::read.xlsx2("MLB_cleaned.xlsx", sheetIndex = 1)
```

\vfill

**Important: Always check if the data was read in correctly!!!**

\vfill
```{r}
head(mlb)
```

## Importing Files: Common issues
\footnotesize

Data looks fine with `head()`, but...

\vfill

```{r}
str(mlb)
```

\vfill

### Recommendation
* Often data types are not read correctly (e.g., numbers might be read as `character` or we want some columns top be factors because they are nominal variables). Variable names have extra dots at the end. These need to be fixed (see next slide).
* There may be issues with different versions of Excel. It is often easier to export the spread sheet from Excel as a CSV file to be read by R.


## Importing Files: Fixing data types
\footnotesize

```{r}
mlb$Height <-  as.numeric(as.character(mlb$Height.inches.))
mlb$Height.inches. <- NULL # remove the column with the old name
mlb$Weight <-  as.numeric(as.character(mlb$Weight.pounds.))
mlb$Weight.pounds. <- NULL
mlb$Age <-  as.numeric(as.character(mlb$Age))
mlb$First.Name <- factor(mlb$First.Name)
mlb$Last.Name <- factor(mlb$Last.Name)
mlb$Team <- factor(mlb$Team)
mlb$Position <- factor(mlb$Position)

str(mlb)
```
\vfill

This looks better. Height, weight and age are now numbers. The name, team and positions are factors.

## Importing Files: Fixing data types
\footnotesize

```{r}
summary(mlb)

mlb_avg <- aggregate(mlb[, c("Age", "Height", "Weight")], 
  by = list(Team = mlb$Team), FUN = mean)
head(mlb_avg)

```

## Exporting Data to Excel

Options:

1. Write data into a CVS file with `write.csv()` and open in Excel.

2. Construct a Excel workbook using multiple calls to `write.xlsx2()`

\vfill

```{r}
library("xlsx")

# create a workbook with two spread sheets
write.xlsx2(mlb, file = "mlb_new.xlsx", 
  sheetName = "MLB Data")

write.xlsx2(mlb_avg, file = "mlb_new.xlsx", 
  sheetName = "MLB Averages", append = TRUE)
```

## Importing an Excel sheet that was saved as CSV.
\small

It is often better to just save the Excel sheet as a CSV file and read it into R.

\vfill

```{r}
mlb <- read.csv(paste0("https://michael.hahsler.net/SMU/",
  "DS_Workshop_Intro_R/examples/MLB_cleaned.csv"), stringsAsFactors = TRUE)
# paste0 is just used because the URL is too long to fit on the slide

str(mlb)
```

\vfill

### Hint
R can directly read from URLs and does an OK job for identifying the correct data types.

# Exercises

## Exercises

1. Read and clean the MLB data set.
3. Select only the players for the team 'ARZ'. Compare the column Team with 'ARZ' and
use subsetting to select the rows.
2. How many players does the team 'ARZ' have in the data set? 
4. What is the weight of the heaviest player of the team 'ARZ' (use a function).
6. What is the average age of all players in the dataset?
7. Add a column called BMI and add the body mass index (https://en.wikipedia.org/wiki/Body_mass_index) for each player.
8. Create a data.frame containing the names, year of birth, month of birth and day of birth as separate columns with the information for 3 people. Make sure the data.frame has column names (see `colnames()`).
9. Write the data.frame to a file in CSV format and check it in Excel.

