---
title: "Intro to R - 3. Functions, Loops and Apply"
author: "Michael Hahsler"
output:
  beamer_presentation:
    colortheme: beaver
    slide_level: 2
    theme: Boadilla
    toc: yes
    highlight: zenburn
    includes:
      in_header: header.tex
    keep_tex:  true
institute: OIT, SMU
subtitle: OIT/SMU Libraries Data Science Workshop Series
fontsize: 9pt   
titlegraphic: WorldChangersRBStackedWebOnlyrgb.png
---


# Functions

```{r, echo = FALSE}
options(width = 70)
options(digits = 2)
# for sampling
set.seed(1234)
```


## Built-in functions

R offeres a wide range of useful functions. Here are some examples:

* Statistics: `min`, `max`, `mean`, `median`, `quantile`, `sd`, `var`, `cor`, `round`
* Sorting: `sort`, `order`, `rev`, `rank` 
* Random numbers: `runif`, `rnorm`, ...

\small
\vfill

```{r}
x <- runif(10)
x

summary(x)

sort(x)
sort(x, decreasing = TRUE)
```

## User defined functions
\small
R is a functional programming language. 
Functions are objects of mode ``function''.

\vfill

```{r}
# defining a function
inc <- function(x) { x + 1 } 
inc
mode(inc)

# calling the function
inc(5)

# functions that use only vectorized operators (e.g., +) 
# are automatically vectorized.
inc(1:10)
```

\vfill

### Note
Functions return the
value of the last expression or can be specified via `return(value)`.

## Named arguments and default values
\small

```{r}
inc <- function(x, by = 1) { x + by } 

inc(5)			# using a default value
inc(1:5, 10)		# using argument order
inc(1:5, by = 10)		# using argument order + names
inc(by = 10, x = 1:5)	# using argument names
inc(matrix(1:4, nrow = 2), 10)
```

\vfill

### Advanced Knowledge
Since functions are regular (first class) objects they can be passed
on as arguments and returned by functions.

## Generic Functions and S3 Objects

**Remember:** Objects have a class that can be seen when calling

```{r}
x <- data.frame(a = 1:2, b = c("A", "B"))
class(x)
```

\vfill

Many functions in R (e.g., `print`, `plot`) look at the supplied object and then choose automatically an appropriate behavior. These functions are called __generic__ functions. 
The __implementations__ have the object type in the name after a dot. 

\vfill

**Example:**
`print` is generic and calls for a data.frame the implementation 
`print.data.frame`.

```{r}
print(x)
```

You can find the manual page using `? print.data.frame`

\vfill

### Hint
When you type a function name in RStudio, it shows you all the implemented methods
in the auto-completion context menu.


# Conditions and Loops

## The if Statement
\small
```{r}
x <- 12
if (x > 10) {	    # result of condition needs length 1
    print("x is > 10") 
} else { 
    print("x is <= 10") 
}

```

\vfill

Using vectors to make decisions

```{r}
x <- c(12, 16, 3)
if (all(x > 10)) print("All values in x are >10")
if (any(x > 10)) print("There is at least one value >10")   

c(FALSE, TRUE, TRUE) |  c(FALSE, TRUE, FALSE) # element-wise OR (see &)
c(FALSE, TRUE, TRUE) || c(FALSE, TRUE, FALSE) # only eval. the 1. elements
```

## The for loop
\small
````{r}
# calculate the sum of the integers 1 to 5
x <- 0
for (i in seq_len(5)) {
    x <- x + i
}
x

# we can also use lists (R recycles values!)
l <- list(a = 2, b = 1:2, c = 4)
x <- 0; for(i in l) { x <- x + i }
x
```

## The while loop
\small

```{r}
# calculate the sum of the integers 1 to 5
x <- 0
i <- 1
while (i <= 5) { x <- x + i; i <- i + 1 }
x

# in R we would rather use a vectorized function
sum(1:5)
```

\vfill
<!-- ###
`next` and `break` and work as expected to skip iterations or end the loop.
-->

### Note
Loops are not frequently used in R since most problems can be 
solved more efficiently using functions  and vectorization.

# apply, lapply, sapply, ...


## apply -- apply functions to a matrix
```{r}
m <- matrix(1:9, nrow = 3)
m

apply(m, MARGIN = 1, sum)	    # apply sum to rows
apply(m, MARGIN = 2, sum)	    # apply sum to cols
# same as
rowSums(m); colSums(m)
```

## lapply/sapply -- apply functions to each element in a lists

```{r}
l <- list(1:3, 6, 7:3)
# apply rev to all elements
lapply(l, FUN = function(x) { rev(x) }) 
# apply automatically "simplifies" the result. Here into a vector.
sapply(l, length)		      
```

# Exercises
## Exercises
1. Create x by `x <- runif(100)`. Write a function with the name 
    `avg_gt` with two formal arguments: a vector x and a value gt.
    The functions computes the average of the values greater than gt in x.
    Write a version with a loop and if and one version without loops and if 
    statements.  

2. Create a list with 5 numeric vectors (lengths and values
    of your choice). Sort all vectors in the list. Hint: see `sort()`.

3. Write a function that computes the smallest value in each column of
    a given matrix.  Create a random $5\times5$ matrix to test the function.

