---
title: "An Extremely Short Introduction to Tidyverse"
author: "[Michael Hahsler](/)"
output:
  rmdformats::readthedown:
    toc_depth: 3
    css: "http://michael.hahsler.net/css/rmarkdown_readthedown.css"
---

<!--
author: "Michael Hahsler"
output:
 html_document:
   toc: true
   toc_float: true
   theme: yeti
   highlight: zenburn
-->

# Introduction

[Tidyverse](https://www.tidyverse.org/) is a collection of packages that makes working with data 
simpler and more consistent.
 
A good overview is given in
the [RStudio Data Transformation Cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-transformation.pdf) and an introduction can be found in the
[Section on Data Wrangling](https://r4ds.had.co.nz/wrangle-intro.html) the free book [R for Data Science](https://r4ds.had.co.nz).

To use
tidyverse you have to install it first with `install.packages("tidyverse")`.
Load tidyverse:

```{r }
library("tidyverse")
```

# Tibbles

A [`tibble`](https://tibble.tidyverse.org/) is a replacement for a `data.frame` that is simpler and often 
faster. 

The iris data comes as a `data.frame`.

```{r }
data("iris")
head(iris)
```

Convert the data into a `tibble`.

```{r }
iris <- as_tibble(iris)
iris
```

tibbles can be used (almost) exactly like data.frames.

# Pipes `%>%` 

Pipes let you compose a sequence of function calls in a more readable way.
The following two lines do the same.

Standard functional form in R using nested functions:

```{r }
print(head(iris))
```

Using pipes makes this more readable as a sequence of operations:

```{r }
iris %>% head() %>% print()
```

The pipe supplies the result of the 
previous function as the **first argument** to the next function.
More information on pipes can be found in the package [magrittr](https://magrittr.tidyverse.org/).

# Data Import with `readr`

Use `read_csv()` (note the `_`). It will guess data types and is faster 
and more versatile compared to the base R read function.

```{r }
mlb <- read_csv("https://michael.hahsler.net/SMU/EMIS3309/R/examples/MLB_cleaned.csv",
                col_types = "ccffddd")
mlb
```

`col_types` can be used to specify the data type for each column (`c` for character, `f` for factor, `d` for double).  

# Data Transformation with `dplyr`

[`dplyr`](https://dplyr.tidyverse.org/) uses pipes to apply a series of functions to data. Functions are similar to SQL's
`SELECT ... FROM ...  WHERE ...  GROUP BY ... ORDER BY ...` syntax. 
:

- Start with a table (FROM in SQL). 
- Pick variables by their names with `select()` (SELECT in SQL).
- Pick observations by their values using `filter()` (WHERE in SQL).
- Reorder the rows using `arrange()` (ORDER BY in SQL).
- Calculate groups summaries using `group_by()` and `summarize()` (GROUP BY and aggregation function in SELECT of SQL).
- Create new variables with functions of existing variables with `mutate()` (a new column in SELECT in SQL).

Variable names (columns) from the data can be directly used in the functions.

Examples:

```{r }
iris %>% select(starts_with("Sepal"))

iris %>% filter(Species == "setosa")

iris %>% arrange(desc(Sepal.Length))

iris %>% group_by(Species) %>% summarize_all(mean)

iris %>% mutate(Sepal.Length_scaled = drop(scale(Sepal.Length)))
```

Complete example:
```{r }
iris %>% filter(Species == "setosa") %>% 
  select(starts_with("Sepal")) %>% 
  arrange(desc(Sepal.Length)) %>% 
  head(3)
```

`dplyr` also provides join functions (e.g., `inner_join()`) to merge multiple tables.

# Tidy data with `tidyr`

Data we get is typically organized in _"wide format"_ where each value has its own row

The iris dataset is in the wide format where each row is a flower and each column
contains the values of a measurement.

```{r }
head(iris)
```

An alternative format is the _"long format."_ Each variable has its own row in this format.

To convert the data into a long format we can use `pivot_longer()`. I first add
a column with the row ID to identify a flower.

```{r }
iris_long <- iris %>% 
  rowid_to_column("ID") %>% 
  pivot_longer(cols = c("Sepal.Length", "Sepal.Width", 
                        "Petal.Length", "Petal.Width")) 

iris_long
```

Note that each flower has now multiple rows, one for each variable.

An application is to use a long format with `ggplot2` facets to produce a histogram for each variable organized by species.

```{r }
ggplot(iris_long, mapping = aes(value)) + 
  geom_histogram() + 
  facet_grid(cols = vars(name), rows = vars(Species))
```

Long format can be converted to a wide format using `pivot_wider()`.

# Strings, Dates, Factors, etc.

Tidyverse provides many packages to work with certain data:

- `tidyr` data cleaning and shaping.
- `forcats` Tools for Working with Categorical Variables (Factors).
- `lubridate` Make Dealing with Dates a Little Easier.
- `stringr` A cohesive set of functions designed to make working with strings as easy as possible. 
- `purr` Apply functions with tidyverse (functional programming).

The best starting point are the [RStudio Cheatsheets](https://www.rstudio.com/resources/cheatsheets/).
