---
title: "Intro to R - 4. Base R Plots"
author: "Michael Hahsler"
institute: OIT, SMU
output:
  beamer_presentation:
    colortheme: beaver
    highlight: zenburn
    includes:
      in_header: header.tex
    slide_level: 2
    theme: Boadilla
    toc: yes
  slidy_presentation:
    highlight: zenburn
subtitle: OIT/SMU Libraries Data Science Workshop Series
fontsize: 9pt
titlegraphic: WorldChangersRBStackedWebOnlyrgb.png
---



# Simple Plots

```{r, echo = FALSE}
options(width = 70)
options(digits = 2)
# for sampling
set.seed(1234)
```


## Introduction
\small

* Ploting is an integral part of R. 

* R plots on devices (e.g., `X11(), quarz(), windows(), pdf(), png()`)

* Plotting commands are divided into three basic groups:
    
    1. **High-level plotting functions** create a new plot on the graphics device, possibly with axes, labels, titles and so on.
    2. **Low-level plotting functions** add more information to an existing plot, such as extra points, lines and labels.
    3. **Interactive graphics functions** allow you interactively add information to, or extract information from, an existing plot, using a pointing device such as a mouse. 

\vfill

We will only discuss "base" graphics here. A very popular alternative is `ggplot`.

# High-level Graphics Functions

## plot
\small

```{r out.height= "60%"}
x <- 1:100
y <- x^2
plot(x, y)
```


## plot
\small

```{r out.height= "60%"}
# use type l = line, line width 3 and red
plot(x, y, type = "l", lwd = 3, col = "red")

# add a vertical dashed (line type 2) line
abline(v=50, lty=2)		 # add a vertical line
```

## Getting help for plot

`? plot`

Shows that plot is a so called generic function. Generic functions
have implementations dor different data types which get "dispatched" 
at call-time.

\vfill

`? plot.default`

This is the default function for plot. 

\vfill

`? par`

Graphical parameters which typically can be passed on as `...` to plot.


## Scatterplot

```{r out.height= "60%"}
mlb <- read.csv(paste0("https://michael.hahsler.net/SMU/",
  "DS_Workshop_Intro_R/examples/MLB_cleaned.csv"))

plot(x = mlb$Height.inches., y = mlb$Weight.pounds., 
  xlab = "Height", ylab = "Weight",
  main = "Weight by Height") 
```

## Scatterplot matrix (pairs plot)

\small

`plot()` is generic which means that it produces different results depending on the type of data.
For example, data frames result in a pairwise scatterplot matrix.

\vfill

```{r out.height= "50%"}
data(iris)
head(iris, n=1)	    # iris is a data.frame
plot(iris[,-5], col= iris[,5]) # use Species (col 5) for color
```

## barplot - Bar Charts

```{r out.height= "50%"}
gpa <- c(Sandra = 3.8, Michael = 4.0, Peter = NA)
gpa.sort <- sort(gpa, na.last = TRUE)
barplot(gpa.sort, ylab = "GPA", xlab = "Student")
```

## barplot - Bar Charts

How many players do the Texas Rangers have for each position?

\tiny
```{r out.height= "60%"}
mlb_tex <- mlb[mlb$Team == "TEX", ]
# use par to make space for labels see ? par and look for mar (plot margin)
oldpar <- par(mar = c(10, 4, 4, 1) + .1)
barplot(sort(table(mlb_tex$Position), decreasing = TRUE), 
  las = 2, ylab = "Count", main = "Texas Rangers: Players per position")
par(oldpar)
```

## hist - Histogram
\small
```{r out.height= "50%"}
hist(mlb$Age, breaks=20)
```

## hist - Histogram with estimated density
\small
```{r out.height= "50%"}
hist(mlb$Age, breaks=20, prob=TRUE)
lines(density(mlb$Age), col="red")
```

## image
\small
volcano is a R data set with elevation measurements of Maunga Whau
on a 10m by 10m grid.

```{r out.height= "50%"}
data(volcano)
dim(volcano)
image(volcano)
```


## Commmon arguments for plot functions

* `add=TRUE`: add to an existing plot?
* `axes=FALSE`: plot axes?
* `log="x"`, `log="y"` or `log="xy"`: Use a logarithmic scale?
* `type="l"`: plot lines instead of points 
* `xlab, ylab`: axis labels
* `main`: figure title
* `sub`: sub-title

# Low-level Graphics Functions
## Some low-level functions

These functions can be used to add elements to a plot.

* `points(x, y)`
* `lines(x, y)`
* `text(x, y, labels, ...)`
* `abline(a, b)` or `abline(h=y)` or `abline(v=x)`
* `polygon(x, y, ...)`
* `legend(x, y, legend, ...)`
* `title(main, sub)`
* `axis(side, ...)`


## Graphical parameter list: par
\footnotesize

R maintains a list of graphics parameters to control
line style, colors, figure arrangement and text justification. 

A separate list of graphics parameters is maintained for each active device.

```{r out.height= "30%"}
oldpar <- par(col = "blue", pch=4, cex = 2)
plot(x, y, axes = FALSE)
axis(1); axis(2)
par(oldpar)	# restore the original parameters
```

Many parameters from `par()` can also be passed to `plot()`.
Try `par()` and `?par`

### Danger
Do not forget to reset `par` to the original settings! 

You can also reset par by closing the plotting device or with the little broom symbol in the Plots tab of RStudio.

## Important parameters in par

* `pch=4`: Plotting symbol (0-25) 
* `lty=2`: Line type  
* `lwd=2`: Line width 
* `col=2`: Color for points, lines, etc. 
* `cex=1.5`: Character expansion (e.g., 50\% larger than default text size) 
* `mai=c(1, 0.5, 0.5, 0)`: Widths of the bottom, left, top and 
right margins, respectively, measured in inches. `mar` is the same, just measured in rows.
* `mfcol, mfrow`: Put multiple plots next to each other.

## Saving a plot as an image

```{r}
png(file="plot.png")	# open device
plot(x, y)
dev.off()		# close device
```

\vfill

\begin{block}{}
Other devices are `jpeg(), tiff(),` `pdf(), postscript(),` `win.metafile()` 
(Windows). 

Use `?Devices` for a complete list.
\end{block}

## Advanced and InteractiveGraphics

The most populare advanced graphics packages are: 

* `ggplot2`: Grammar of graphics. Produces elegant visualizations 
(see http://ggplot2.org/).  
* `grid`: Advanced graphics can be programmed using flexible
low level ploting functions (viewports, different coordinate systems
and units, lines, points, text, etc.) See also package `lattice`.

\vfill

Interactive Graphics are available via several extension packages.
Here are some examples:

https://www.r-graph-gallery.com/interactive-charts.html





# Exercises
## Exercises

\begin{enumerate}
\item Plot a `sin(x)/x`. Hint: Trigonometric functions in R use angles in
radians (see `sin`)
\item 
The ``cars'' data set gives the speed of cars and the distances taken to stop.
     Note that the data were recorded in the 1920s.
Plot the ``cars'' data set as a scatter plot. Plot all 
data points with distances taken to stop greater than 80 in red. 
\item Plot histograms for speed and dist in ``cars''.
\end{enumerate}

