---
title: "MLB Dashboard Example"
author: "Michael Hahsler"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    social: menu
    source_code: embed
---

```{r setup, include=FALSE}
library(flexdashboard)
mlb <- read.csv("https://michael.hahsler.net/SMU/DS_Workshop_Intro_R/examples/MLB_cleaned.csv")
```

MLB Data
============================

Column {data-width=650}
-----------------------------------------------------------------------

### MLB Roster

```{r}
DT::datatable(mlb)
```


Column {data-width=350}
-----------------------------------------------------------------------

### Age Distribution

```{r}
plotly::plot_ly(mlb,  x = ~Age, type = "histogram")
```

### Weight by height

```{r}
plotly::plot_ly(mlb,  x = ~Height.inches. , y = ~Weight.pounds., text = ~paste(First.Name, Last.Name, Team))
```

About
===============================

This dashboard was developed for the workshop [Introduction to Programming with R.](https://michael.hahsler.net/SMU/DS_Workshop_Intro_R/)

More information on flexdashboards can be found at <https://rmarkdown.rstudio.com/flexdashboard/>.


__How to share a dashboard?__

Flexdashboards are self-contained in a single html file (including data and figures). You share it with others by sending the html file and they can open it in any web browser.
